\documentclass{beamer}

\usepackage{color}
\definecolor{keywordcolor}{rgb}{0.7, 0.1, 0.1}   % red
\definecolor{tacticcolor}{rgb}{0.0, 0.1, 0.6}    % blue
\definecolor{commentcolor}{rgb}{0.4, 0.4, 0.4}   % grey
\definecolor{symbolcolor}{rgb}{0.0, 0.1, 0.6}    % blue
\definecolor{sortcolor}{rgb}{0.1, 0.5, 0.1}      % green
\definecolor{attributecolor}{rgb}{0.7, 0.1, 0.1} % red

\usepackage{listings}
\def\lstlanguagefiles{lstlean.tex}
\lstset{language=lean}

\usepackage[T1]{fontenc}
\usepackage{amsmath, amssymb, upgreek}
\usepackage{stmaryrd}
\usepackage{bussproofs}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{unixode}

\newcommand{\ttt}[1]{\texttt{#1}}
\DeclareTextFontCommand{\emph}{\bfseries}

\newcommand{\bN}{\mathbb{N}}
\newcommand{\bZ}{\mathbb{Z}}
\newcommand{\bR}{\mathbb{R}}
\newcommand{\Type}{\ttt{Type}}
\newcommand{\Prop}{\ttt{Prop}}
\newcommand{\fst}{\mathrm{fst}}
\newcommand{\snd}{\mathrm{snd}}
\newcommand{\letin}[3]{\mathrm{let}\ #1\ := #2\ \mathrm{in}\ #3}
\newcommand{\pair}[2]{\langle #1,\ #2\rangle}
\newcommand{\uPi}{\Uppi}
\newcommand{\lam}{\lambda}
\newcommand{\lx}{\lam x}
\newcommand{\ra}{\rightarrow}
\renewcommand{\th}{\vdash}
\newcommand{\sep}{\ \mid\ }
\newcommand{\subst}[2]{[#2/#1]}
\newcommand{\ueq}{\cong}
\newcommand{\I}[1]{\llbracket #1\rrbracket}
\newcommand{\mvar}{?\!}
\newcommand{\lean}{Lean}

\title{Interactive Theorem Proving with Lean}

\author{Cody Roux\\ Draper Labs}

\begin{document}

\begin{frame}
  \maketitle
\end{frame}

\section{Overview}

\begin{frame}{What is theorem proving?}
  The field of Theorem Proving investigates expressing and proving
  \emph{mathematical theorems} with the help of computers

\end{frame}

\begin{frame}{What is it good for?}
  Any mathematical theorem is game, but one of the prime applications
  is verification of \emph{hardware and software applications.}
\end{frame}

\begin{frame}{What is it good for... in CS?}
  \begin{center}
    ``If you're dealing with life and death, consider proving some key
    algorithms. But in the real world, you arent gonna need it.'' --
    user RonJeffries, c2.com
  \end{center}
  \begin{center}
    ``[T]he absence of continuity, the inevitability of change, and
    the complexity of specification of significantly many real
    programs make the formal verification process difficult to justify
    and manage.'' -- De Millo, Lipton and Perlis, {\it Social
      Processes and Proofs of Theorems and Programs (1979)}
  \end{center}
  \begin{center}
    ``Unfortunately, there is a wealth of evidence that automated
    verifying systems are out of the question'' -- {\it ibid}
  \end{center}
  \begin{center}
    ``The amount of formal reasoning required to prove even the
    smallest programs is beyond the capacity of most of us.'' --
    anonymous user, c2.com
  \end{center}
\end{frame}

\begin{frame}{What is it good for... in CS?}
  But...
  \begin{center}
    ``The striking thing about our CompCert results is that the
    middle-end bugs we found in all other compilers are absent. As of
    early 2011, the under-development version of CompCert is the only
    compiler we have tested for which Csmith cannot find
    wrong-code errors. This is not for lack of trying: we have devoted
    about six CPU-years to the task. The apparent unbreakability of
    CompCert supports a strong argument that developing compiler
    optimizations within a proof framework, where safety checks are
    explicit and machine-checked, has tangible benefits for compiler
    users.'' -- Yang {\it et al.}, {\it Finding and Understanding Bugs
      in C Compilers}
  \end{center}
\end{frame}

\begin{frame}{Some Milestones}
  In Math:
  \begin{itemize}
  \item The Four Color Theorem (2005, Coq) \emph{every map can be colored with at most four colors}
  \item The Prime Number Theorem (2005, Isabelle) \emph{the prime numbers grow like $n/\log(n)$}
  \item The Feit-Thompson Theorem (2012, Coq) \emph{every group of odd order is solvable}
  \item The Kepler Conjecture (2014, HOL) \emph{we know how to pack oranges}
  \end{itemize}
  In CS:
  \begin{itemize}
  \item Verification of the driverless Metro 14 line in Paris (1998, B-method)
  \item Verification of the CompCert C compiler (2009, Coq)
  \item Verification of the seL4 microkernel (2014, Isabelle)
  \item Many more!
  \end{itemize}
\end{frame}

\begin{frame}{But what about me?}
  Alright, enough sales pitch!\bigskip

  How can you get started with theorem proving?
\end{frame}

\begin{frame}{Lean}
  Ok, some more sales pitch:\bigskip

  In this talk, I'll be using a new theorem prover called Lean, developed by
  \begin{itemize}
  \item Leo de Moura, at Microsoft Research
  \item Jeremy Avigad, Soonho Kong, Floris van Doorn at CMU
  \item And many others!
  \end{itemize}

  Can be found at \url{http://leanprover.github.io}
\end{frame}
  
\section{The Core Components}

\begin{frame}{The core components}
  An Interactive Theorem Prover (ITP) is
  \begin{itemize}
  \item
    \begin{block}{A theorem prover:}
      It needs to be able to \emph{verify} fully detailed formal
      proofs of theorems. We call this the \emph{Kernel}
    \end{block}
  \item
    \begin{block}{Interactive}
      It needs to help the user provide such proofs. This is called
      the \emph{Elaborator} and \emph{Proof Engine}.
    \end{block}
  \end{itemize}
\end{frame}

\begin{frame}{The core components}
  Two silly examples:
  \begin{block}{The ideal prover}
    \begin{itemize}
    \item The kernel: Some implementation of a powerful logic that can express all abstractions.
    \item The proof engine: Just a \ttt{Prove} button!
    \end{itemize}
  \end{block}
  \begin{block}{The simplest prover}
    \begin{itemize}
    \item The kernel: some implementation of a simple logic, like First Order Logic.
    \item The engine: just a proof parser.
    \end{itemize}
  \end{block}
  \begin{block}{Reality}
    Somewhere in between.
  \end{block}
\end{frame}


\begin{frame}{The Kernel}
  The kernel is the most important \emph{trusted component}. We need
  \begin{itemize}
  \item A simple but expressive logic
  \item A trustworthy implementation
  \end{itemize}
  The ideal implementation:

  \begin{center}
    \begin{tikzpicture}
      
      
      \node (1) at (0,0) {Simple};
      \node (2) at (2,2) {Small};
      \node (3) at (4,0) {Fast};
      
      \draw[-] (1) -- (2) -- (3) -- (1);
    \end{tikzpicture}
  \end{center}


  Choose 2!
\end{frame}

\begin{frame}{The logic}
  We want simple (trustworthy, easy to implement) but powerful enough
  to formalize anything.
  \begin{block}{We use}
    \begin{itemize}
    \item A theory of \emph{dependent types} that unifies \emph{function
      spaces, universal quantifiers and polymorphism}
    \item An impredicative type of propositions
    \item A hierarchy of \emph{universes}
    \item Inductive families (GADTs in Haskell)
    \item Quotient types (useful to build $\bZ$)
    \end{itemize}
  \end{block}
  This is similar to the logic of Coq.
\end{frame}

\begin{frame}{The implementation}
  \begin{itemize}
  \item Simple
  \item Very Fast
  \item In C++
  \item $\sim$ 6000 lines of code
  \end{itemize}

  There is an alternate implementation of the kernel, in Haskell (a
  couple thousand lines).
\end{frame}

\section{A specification example}

\begin{frame}{A specification example}
  Let's work through an example to demonstrate the specification
  language:
  \begin{block}{Let's formalize the statement}
    There is an infinite number of primes.
  \end{block}
\end{frame}

\begin{frame}[fragile]{A specification example}
  First observation: logical connectives can be represented by
  inductive types
  \begin{lstlisting}
    inductive and (a b : Prop) : Prop :=
      intro : a → b → and a b

    inductive or (a b : Prop) : Prop :=
    | inl {} : a → or a b
    | inr {} : b → or a b

    inductive ⊤ : Prop := trivial : true

    inductive ⊥ : Prop

    definition ¬ (a : Prop) : Prop := a → ⊥

  \end{lstlisting}
  
\end{frame}
\begin{frame}[fragile]{A specification example}
  Even existence and equality
  \begin{lstlisting}
inductive Exists {A : Type} (P : A → Prop) : Prop :=
    intro : ∀ (a : A), P a → Exists P

    

inductive eq {A : Type} (a : A) : A → Prop :=
    refl : eq a a
  \end{lstlisting}
  
\end{frame}

\begin{frame}[fragile]{A specification example}
  Natural numbers are also specified by induction
  \begin{lstlisting}
    inductive ℕ :=
    | zero : ℕ
    | succ : ℕ → ℕ
  \end{lstlisting}
  And operations are defined by recursion
  \begin{lstlisting}
    definition add : ℕ → ℕ → ℕ
    | add 0 m := m
    | add (succ n') m := succ (add n' m)

    definition mul : ℕ → ℕ → ℕ
    | mul 0 m := m
    | mul (succ n') m := add m (mul n' m)
  \end{lstlisting}
\end{frame}

\begin{frame}[fragile]{A specification example}
  with a few notations
  \begin{lstlisting}
    notation n `+` m := add n m
    notation n `*` m := mul n m
    ...
  \end{lstlisting}
  \begin{lstlisting}
    definition le (n m : ℕ) := ∃ k, n + k = m
    
    definition dvd (n m : ℕ) := ∃ k, n * k = m
    
    definition prime (p : ℕ) :=
    ∀ d, 2 ≤ p ∧ (d ∣ p → d = 1 ∨ d = p)
  \end{lstlisting}
  We can then express (and prove)
  
  \begin{lstlisting}
    theorem inf_primes := ∀ n, ∃ p, n ≤ p ∧ prime p
  \end{lstlisting}
\end{frame}

\section{Elaboration}

\begin{frame}[fragile]{The elaborator}
  To have any hope of being able to actually specify and prove
  anything, we need
  \begin{enumerate}
  \item a synthesizer for implicit arguments
    \begin{lstlisting}
 eq.subst : ∀ {A : Type}{a b : A}{P : A → Prop},
      a = b → P a → P b
    \end{lstlisting}
    we need to \emph{synthesize} (guess) values for $A, a, b$ and $P$!
    
  \item a language for proofs
  \end{enumerate}
\end{frame}

\begin{frame}[fragile]{The elaborator}
  This problem is non-obvious:
    \begin{lstlisting}
 eq.subst : ∀ {A : Type}{a b : A}{P : A → Prop},
      a = b → P a → P b
    \end{lstlisting}

    Say $P\ a\ \equiv\ a + a = 2$. Then we can have
    \[P\ \equiv\ \lambda x.\ x+x = 2\]
    or
    \[P\ \equiv\ \lambda x.\ x+a = 2\]
    \[P\ \equiv\ \lambda x.\ a+a = 2\]
    we need to account and search for \emph{multiple solutions}.
\end{frame}

\section{Proofs}

\begin{frame}{The proof language}
  How to design an effective high-level proof language is one of the
  major open questions in ITP research.\bigskip

  Right now there are 2 main approaches:
  \begin{itemize}
  \item A ``stack based'' \emph{tactic language}, which operates by
    transforming the goal in a series of composable steps. Very
    powerful, but hard to read
  \item A declarative ``isar-like'' language, which is more verbose
    but easier to read and understand
  \end{itemize}
  
  On top of this, we need some powerful proof automation to help with
  the simpler intermediate steps.
\end{frame}

\begin{frame}{Let's see some code!}
  Let's work through some simple examples.\bigskip

  Now is a good time for
  \begin{center}
    {\huge \bf Questions}
  \end{center}
\end{frame}

\section{Technical features of Lean}

\begin{frame}[fragile]{A little more depth...}
  Some additional features:
  \begin{itemize}
  \item Implicit coercions
  \item Type Classes and ``structures''
  \item Complex notations (infix, postfix, mixfix,...)
  \item Namespace management: local scopes, notations, coercions
  \item Lua bindings
  \item Javascript front end
  \item Quotient types
  \item A HoTT mode
  \end{itemize}
\end{frame}

\begin{frame}[fragile]{Coercions}
  An important part of elaboration is the ability to insert
  \emph{implicit casts}
\small{
  \begin{lstlisting}
variable {A : Type}
definition to_list [coercion] :
               ∀ {n : nat}, vector A n → list A
| to_list 0 nil := nil
| to_list (n+1) (a::v) := a::(to_list v)
variable f : list nat → Prop
variable v : vector nat 10
check f v -- f (@to_list nat 10 v)
  \end{lstlisting}
}
\end{frame}

\begin{frame}[fragile]{Classes and structures}
  The notion of \emph{dependent record} is a really natural way of
  representing \emph{mathematical structures}
\small{
  \begin{lstlisting}
 structure group (A : Type) :=
 (mul : A → A → A)
 (mul_assoc : ∀ a b c, mul (mul a b) c = mul a (mul b c))
 (...)
  \end{lstlisting}}
But it's better to use classes and inheritance for modularity
\small{
  \begin{lstlisting}
 structure has_mul [class] (A : Type) :=
    (mul : A → A → A)
 structure semigroup [class] (A : Type) extends has_mul A :=
 (mul_assoc : ∀ a b c, mul (mul a b) c = mul a (mul b c))
  ...
  \end{lstlisting}}

\end{frame}

\begin{frame}[fragile]{Decidable Type Class}
    An element of \Prop is said to be decidable if we can decide
    whether it is true or false.

    \begin{lstlisting}
      inductive decidable [class] (p : Prop) : Type :=
      | inl : p → decidable p
      | inr : ¬p → decidable p
    \end{lstlisting}

    Having an element \ttt{t : decidable p} is stronger than having an element \ttt{t : p ∨ ¬p}
    
    The expression \ttt{if c then t else e} contains an implicit argument \ttt{[d : decidable c]}.
    If Hilbert's choice is imported, then all propositions are decidable (smooth transition to classical reasoning).
\end{frame}

\begin{frame}[fragile]{Quotient types}
  We support \emph{quotient types}
  \begin{itemize}
  \item For any type $A$ that supports an equivalence relation $R$ we
    can form a type $A/R$
  \item $A/R$ has the \emph{universal property}: there is a function
    \[f:A/R \rightarrow B\]
    exactly when there is a $g:A\rightarrow B$ such that
    \[\forall x y,\ x\ R\ y\ \Rightarrow g(x) = g(y)\]
  \end{itemize}
\end{frame}

\begin{frame}[fragile]{HoTT}
  There has been a lot of buzz around the recent \emph{synthetic
    approach to homotopy theory} based on type theory.\bigskip

  There is a mode in Lean designed for this. Some features
  \begin{itemize}
  \item No impredicative type \Prop
  \item A special case of \emph{higher inductive types} ($n$-truncations)
  \end{itemize}
\end{frame}


\begin{frame}[fragile]{Libraries}
  A theorem prover is only as good as its libraries. Lean already has several
  \begin{itemize}
  \item \ttt{init}: the basic definitions of logical connectives,
    quantifiers, equality and $\bN$
  \item \ttt{data}: All the programmers favorites: lists, ints, bools,
    finite sets and real numbers
  \item \ttt{algebra}: the hierarchy of algebraic structures, including groups, rings and fields
  \item \ttt{theories}: more fleshed out logical theories, with group
    theory (up to Sylow's theorems!), combinatorics, basic analysis
    and number theory
  \end{itemize}
\end{frame}


\begin{frame}[fragile]{What about Coq?}
  The most obviously similar ITP in the same design space is Coq.
  \begin{block}{The major differences}
    \begin{itemize}
    \item Similar core theories: dependent types, impredicative, universes
    \item Some additional features in Lean: Quotient types, HoTT mode
    \item Lean has recursors instead of built-in pattern matching
      (harder to get wrong)
    \item Less automation (for now)
    \item A mechanism for trust: you can call un-trusted decision
      procedures, which can be run in ``safe mode'': the proof is then
      completely re-built
    \end{itemize}
  \end{block}
 
\end{frame}


%% \begin{frame}[fragile]{}
%% \end{frame}

\end{document}


